---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This is my submission for Reproducible Research Course Project.

+ To read more information view the [ReadMe](https://github.com/dbwebhead/reproducible-research/blob/master/README.md) on GitHub.

+ The data for the assignment can be downloaded [here](https://github.com/dbwebhead/reproducible-research/blob/master/activity.zip). 

## Loading and preprocessing the data
*Show any code that is needed to:*  
*1. Load the data (i.e. read.csv())*  
*2. Process/transform the data (if necessary) into a format suitable for your analysis*  

```{r readtable, echo=TRUE, message=FALSE, warning=FALSE}
# import the activity data file
activity_df <- read.csv("activity.csv")

## Install necessary packages
if (!require("ggplot2")) {
        install.packages("ggplot2", dependencies = TRUE)
}	
if (!require("dplyr")) {
        install.packages("dplyr", dependencies = TRUE)
}

# invoke libraries
library(ggplot2)
library(dplyr)
Sys.setlocale("LC_TIME", "English")
```

```{r, echo=TRUE}
# some information about the variables
str(activity_df)
```

As shown, the variables included in this data set are:  

1. **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)  

2. **date**: The date on which the measurement was taken in YYYY-MM-DD format  

3. **interval**: Identifier for the 5-minute interval in which measurement was taken  

## What is mean total number of steps taken per day?

*For this part of the assignment, you can ignore the missing values in the dataset.*  
*1. Calculate the total number of steps taken per day*  
*2. Make a histogram of the total number of steps taken each day*  
*3. Calculate and report the mean and median total number of steps taken per day* 

**1. Number of steps per day:**

```{r stepsperday, echo=TRUE}
# create and print number of steps per day
StepsPerDay <- aggregate(activity_df$steps, list(activity_df$date), 
                         FUN=sum, na.rm=TRUE)
colnames(StepsPerDay) <- c("Date", "Steps")
StepsPerDay
```


**2. Histogram of the total number of steps taken each day:**  

+ Below is a histogram of the aggregate number of steps taken per day.  
+ On the histogram, a blue and a red vertical line is drawn to show the mean and median value of the data, respectively.

```{r histogram1, echo=TRUE, fig.width=10, warning=FALSE}
# draw the histogram
ggplot(StepsPerDay, aes(Steps)) +
	geom_histogram(boundary=0, binwidth=2500, col="darkblue", fill="lightblue") + 
	ggtitle("Histogram of Steps per Day") + 
	xlab("Steps") + 
	ylab("Frequency") + 
	theme(plot.title = element_text(face="bold", size=12)) + 
        geom_vline(xintercept=mean(StepsPerDay$Steps, na.rm=TRUE), size=1, color="blue", alpha=.5) +
        geom_vline(xintercept=median(StepsPerDay$Steps, na.rm=TRUE), size=1, color="red", alpha=.5) +
	scale_x_continuous(breaks=seq(0,25000,2500)) + 
	scale_y_continuous(breaks=seq(0,18,2))
```


**3. Mean and median of total number of steps taken per day:**

```{r calcavg1, echo=TRUE}
# calculate mean (ignoring missing values)
stepMean <- mean(StepsPerDay$Steps, na.rm=TRUE)
print(paste("The Mean number of steps per day is", round(stepMean,1)))

# calculate median (ignoring missing values)
stepMedian <- median(StepsPerDay$Steps, na.rm=TRUE)
print(paste("The Median number of steps per day is", round(stepMedian,1)))
```


## What is the average daily activity pattern?  
*1. Make a time series plot (i.e., type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).  2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?* 

+ With the time series plot below, it is evident that the maximum number of steps on average across all the days appears to be a little over 200.  This occurs approximately at the 8-minute mark.  
<br>
**1. Time series plot of the 5 minute interval (x) and average number of steps taken averaged across all days (y):**

```{r timeplot1, echo=TRUE, fig.width=10, warning=FALSE}
# create table with steps per time
StepsPerTime <- aggregate(steps~interval, data=activity_df, FUN=mean, na.action=na.omit)

# set variable time to make the graph axis more coherent
StepsPerTime$time <- StepsPerTime$interval/100

# draw the line plot
ggplot(StepsPerTime, aes(time, steps)) + 
        geom_line(col="dark green") + 
        ggtitle("Average Steps per Time Interval") + 
        xlab("Time") + ylab("Avg Number of Steps") + 
        theme(plot.title = element_text(face="bold", size=12))
```

**2. Display the 5-minute interval (on average across all the days) with the maximum number of steps**

```{r fiveminint, echo=TRUE, fig.width=10, warning=FALSE}
# table for dplyr
steps_df <- tbl_df(StepsPerTime)

# find the column
steps_df %>% select(time, steps) %>% filter(steps==max(steps_df$steps))
```
<br>

## Imputing missing values

+ *Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.**  
**1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)*  
*2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.*  
*3. Create a new dataset that is equal to the original dataset but with the missing data filled in.*  
*4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.*  
+ *Do these values differ from the estimates from the first part of the assignment?*  
+ *What is the impact of imputing missing data on the estimates of the total daily number of steps?* 

<br>
**1. Determine total number of missing values in the dataset:**

```{r missingval,  echo=TRUE}
# table for dplyr
activity_tbl <- tbl_df(activity_df)

head(activity_tbl)

# find the column
activity_tbl %>% 
   filter(is.na(steps)) %>% 
   summarize(missing_values = n())
```

<br>
**2. Replace missing values:**  

+ The rounded values of the average 5-minute interval is used to replace the NA values.  
+ *CompleteSteps* is the new column without missing values.

```{r imputena,  echo=TRUE}
# values with NA are imputed in a new column
activity_df$CompleteSteps <- ifelse(is.na(activity_df$steps), round(StepsPerTime$steps[match(activity_df$interval, StepsPerTime$interval)],0), activity_df$steps)
```

<br>
**3. Create a new dataset that is equal to the original dataset but with the missing data filled in:**  

+ The first ten values of the new data set are shown below. 

```{r dispimpdf, echo=TRUE}
# create new imputed dataset
activityFull_df <- data.frame(steps=activity_df$CompleteSteps, 
                              interval=activity_df$interval, date=activity_df$date)

# display the first ten rows of the new dataset
head(activityFull_df, n=10)
```
<br>
**4A. Histogram of the total number of steps taken each day (with missing data filled in):** 

```{r histogram2, echo=TRUE, fig.width=10, warning=FALSE}
# prepare data
StepsPerDayFull <- aggregate(activityFull_df$steps, list(activityFull_df$date), FUN=sum)
colnames(StepsPerDayFull) <- c("Date", "Steps")

# draw the histogram
ggplot(StepsPerDayFull, aes(Steps)) +
	geom_histogram(boundary=0, binwidth=2500, col="darkblue", fill="lightblue") + 
	ggtitle("Histogram of Steps per Day (Imputed)") + 
	xlab("Steps") +
	ylab("Frequency") + 
        geom_vline(xintercept=mean(StepsPerDayFull$Steps, na.rm=TRUE), size=1, color="blue", alpha=.5) +
        geom_vline(xintercept=median(StepsPerDayFull$Steps, na.rm=TRUE), size=1, color="red", alpha=.5) +
	theme(plot.title = element_text(face="bold", size=12)) +
	scale_x_continuous(breaks=seq(0,25000,2500)) + 
	scale_y_continuous(breaks=seq(0,26,2))
```

**4B. Calculate and report the mean and median total number of steps taken per day:**  

+ *Do these values differ from the estimates from the first part of the assignment?*    
+ *What is the impact of imputing missing data on the estimates of the total daily number of steps?* 

```{r calcavg2, echo=TRUE}
# calculate mean (missing values imputed)
stepMeanImp <- mean(StepsPerDayFull$Steps)
print(paste("The imputed Mean number of steps per day is", round(stepMeanImp,2)))

# calculate median (missing values imputed)
stepMedianImp <- median(StepsPerDayFull$Steps)
print(paste("The imputed Median number of steps per day is", round(stepMedianImp,2)))
```

+ Evidently, imputing the missing data has minimal impact on the mean and the median of the total daily number of steps.  
+ Watching the histogram we note that the only bin that has changed is the interval between 10000 and 12500 steps, which has grown from a frequency of 18 to a frequency of 26.  
+ Also note that different methods for replace missing values may cause different results.

<br>

## Are there differences in activity patterns between weekdays and weekends?  
+ *For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.*  
*1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.*  
*2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).* 

**1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day:**  

+ *DayType* is the new column indicating if the day is a weekday day or weekend day.  
+ The first ten values of the new table are shown below:

```{r actdiff, echo=TRUE}
# create new date variable in yyyy-mm-dd format 
activityFull_df$RealDate <- as.Date(activityFull_df$date, format = "%Y-%m-%d")

# create a weekday variable
activityFull_df$weekday <- weekdays(activityFull_df$RealDate)

# create a new day type variable (indicating weekday or weekend day)
activityFull_df$DayType <- ifelse(activityFull_df$weekday=='Saturday' | 
                        activityFull_df$weekday=='Sunday', 'weekend','weekday')

# display first ten values
head(activityFull_df, n=10)
```
<br>
**2. Create two time series plot of the 5-minute interval (x) and the average number of steps taken averaged across weekday days or weekend days (y):** 

```{r timeplot2, echo=TRUE, fig.width=10, warning=FALSE}
# create table with steps per time across weekday days or weekend days
StepsPerTime_dt <- aggregate(steps~interval + DayType, data=activityFull_df, FUN=mean, na.action=na.omit)

# set variable time (for the graph axis)
StepsPerTime_dt$time <- StepsPerTime$interval/100

# draw the line plot
ggplot(StepsPerTime_dt, aes(time, steps)) +
        geom_line(col="darkred") +
        ggtitle("Average Steps per Time Interval: Weekdays vs. Weekends") + 
        xlab("Time")+ylab("Steps") + 
        theme(plot.title = element_text(face="bold", size=12)) + 
        facet_grid(DayType ~ .)
```
